% LatticeFit: Journal of Open Research Software submission
% Ubiquity Press — free, peer-reviewed, DOI, indexed
% Target: ~2000 words, article class

\documentclass[10pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{authblk}
\usepackage{microtype}

\hypersetup{colorlinks=true, linkcolor=blue, urlcolor=blue, citecolor=blue}

\lstset{
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{gray!8},
  frame=single,
  breaklines=true,
  language=Python,
  keywordstyle=\color{blue!70},
  commentstyle=\color{gray},
  stringstyle=\color{brown},
}

\newcommand{\doi}[1]{\href{https://doi.org/#1}{\texttt{doi:#1}}}

% ── Title ──────────────────────────────────────────────────────────
\title{\textbf{LatticeFit}: A Python Package for Detecting and
Validating Discrete Multiplicative Structure in Empirical Data}

\author[1]{Marvin L. Gentry}
\affil[1]{Independent Researcher, Seattle, WA, USA.
  \texttt{drmlgentry@protonmail.com}}
\date{March 2026}

\begin{document}
\maketitle

% ── Abstract ───────────────────────────────────────────────────────
\begin{abstract}
\texttt{LatticeFit} is an open-source Python package for detecting
discrete multiplicative lattice structure in positive-valued empirical
datasets. Given a dataset, it fits the model $x \approx A \cdot r^{k/d}$
(anchor $A$, base $r$, denominator $d$, integer $k$), computes a
root-mean-square logarithmic residual, and assesses statistical
significance via log-uniform and sector-anchor null hypothesis tests.
An auto-discovery mode searches over candidate bases and denominators
using the Akaike Information Criterion. The package is validated
across 18 real-world datasets spanning particle physics, geophysics,
biology, finance, social media, and e-commerce, recovering known
scaling laws automatically and returning correct negatives for
structureless data. \texttt{LatticeFit} is MIT-licensed and available
at \url{https://github.com/drmlgentry/latticefit}.
\end{abstract}

\textbf{Keywords:} multiplicative scaling; geometric lattice;
statistical testing; null hypothesis; Python; data analysis

\vspace{1em}

% ── Metadata table (required by JORS) ─────────────────────────────
\begin{table}[h!]
\centering
\caption{Software metadata}
\label{tab:meta}
\small
\begin{tabular}{llp{7cm}}
\toprule
\textbf{Nr.} & \textbf{Field} & \textbf{Value} \\
\midrule
C1 & Current code version & v0.2.0 \\
C2 & Permanent link to code/repository &
     \url{https://github.com/drmlgentry/latticefit} \\
C3 & Permanent link to reproducible capsule & N/A \\
C4 & Legal code license & MIT \\
C5 & Code versioning system used & git \\
C6 & Software code languages, tools, and services used &
     Python ($\geq$3.10); numpy, pandas, scipy, matplotlib \\
C7 & Compilation requirements, operating environments \& dependencies &
     \texttt{pip install latticefit} \\
C8 & If available, link to developer documentation/manual &
     \url{https://github.com/drmlgentry/latticefit/blob/main/README.md} \\
C9 & Support email for questions & drmlgentry@protonmail.com \\
\bottomrule
\end{tabular}
\end{table}

% ── 1. Overview ────────────────────────────────────────────────────
\section{Overview}

Discrete multiplicative structure --- where positive-valued observations
cluster near a geometric lattice $x \approx A \cdot r^{k/d}$ for
integer $k$ --- arises in diverse empirical settings. Known examples
include the equal-tempered musical scale ($r = 2^{1/12}$, exact),
earthquake energy release following the Gutenberg--Richter law
\cite{GutenbergRichter1944} (equivalent to base~$\sqrt{2}$), and
Standard Model fermion masses near a golden-ratio lattice
\cite{Gentry2026masses}. Despite this breadth, no general-purpose
statistical tool has existed for detecting, quantifying, and validating
such patterns across arbitrary datasets.

\texttt{LatticeFit} fills this gap. Unlike tools for continuous power-law
fitting (\texttt{powerlaw}~\cite{powerlaw}) or general curve fitting
(\texttt{lmfit}~\cite{lmfit}), \texttt{LatticeFit} specifically addresses
the problem of discrete geometric lattice clustering in the log domain.
Its core contributions are: (1)~a rigorous RMS-based fit metric with a
natural maximum, (2)~two complementary null hypothesis tests, (3)~AIC/BIC
model selection over candidate bases and denominators, and (4)~a
publication-bundle generator for reproducible reporting.

% ── 2. Functional description ──────────────────────────────────────
\section{Functional description}

The core algorithm assigns each observation $x_i$ a nearest-integer
lattice label $k_i = \operatorname{round}(d \cdot \log_r(x_i/A))$,
computes the logarithmic residual $\delta_i = |\log_r(x_i/A) - k_i/d|$,
and returns the RMS $= \sqrt{n^{-1}\sum_i \delta_i^2}$.
The maximum possible RMS for denominator~$d$ is $0.5/d$, providing
a scale-free quality measure.

Two null tests assess significance. The \emph{log-uniform null}
draws $N$ random datasets uniformly in $[\log x_{\min}, \log x_{\max}]$;
the $p$-value is the fraction achieving RMS $\leq$ observed.
The \emph{sector-anchor null} holds within-group ratios fixed while
randomising group anchors, isolating inter-sector alignment as the
signal of interest. Auto-discovery searches bases
$\{\varphi, \sqrt{2}, 2, e, 10, \ldots\}$ and denominators
$\{1,2,3,4,6,8,12\}$, ranked by AIC.

% ── 3. Software description ────────────────────────────────────────
\section{Software description}

\texttt{LatticeFit} is implemented in pure Python with dependencies
\texttt{numpy}~\cite{numpy}, \texttt{pandas}~\cite{pandas},
\texttt{scipy}~\cite{scipy}, and optionally \texttt{matplotlib}~\cite{matplotlib}.

\noindent\textbf{Python API:}
\begin{lstlisting}
from latticefit import fit, discover
from latticefit.stats import log_uniform_null
from latticefit.bootstrap import bootstrap_ci

# Fit golden-ratio lattice to SM fermion masses
masses = [5.11e-4, 0.106, 1.777, 0.00216, 1.275, 172.76,
          0.00467, 0.0934, 4.18]
result = fit(masses, anchor=5.11e-4, base=1.618034, denom=4)
null   = log_uniform_null(result, n_trials=50000)
boot   = bootstrap_ci(result, n_bootstrap=2000)
print(f"RMS={result.rms:.4f}  p={null.p_value:.3f}")
# RMS=0.0688  p=0.074

# Auto-discover best lattice
top = discover(masses, top_n=5)
\end{lstlisting}

\noindent\textbf{Command-line interface:}
\begin{lstlisting}[language=bash]
latticefit masses.csv --auto --null 50000
latticefit masses.csv --bundle output/ --bootstrap 2000
latticefit --demo earthquakes   # recover Gutenberg-Richter
latticefit --demo musical_notes # exact recovery (RMS~0)
\end{lstlisting}

The publication bundle command produces a \texttt{results.json},
\texttt{table.tex}, \texttt{methods.txt}, and \texttt{reproduce.py}
from a single invocation, enabling fully reproducible reporting.

% ── 4. Illustrative examples ────────────────────────────────────────
\section{Illustrative examples}

\subsection{Exact recovery: equal-tempered musical scale}

The 13 frequencies $f_n = 440 \cdot 2^{n/12}$~Hz form a geometric
lattice by construction. \texttt{LatticeFit} recovers
$r = 2^{1/12}$, $d = 1$ with RMS~$\approx 0$ and $p < 0.001$,
validating the algorithm.

\subsection{Known law recovery: Gutenberg--Richter}

Applied to 2,054 USGS M4.5+ earthquake energies ($E \propto 10^{1.5M}$),
auto-discovery returns base~$\sqrt{2}$, $d = 2$, RMS~$= 0.011$,
$p < 0.001$ --- automatically recovering the Gutenberg--Richter
scaling law without prior domain knowledge.

\subsection{Novel finding: YouTube engagement}

Applied to 16,121 deduplicated YouTube Brazil trending videos,
like counts yield best base~$\sqrt{2}$, $d = 12$, $p = 0.002$
($n = 16{,}121$, 6.3 decades). Comment counts give $p < 0.001$
for both $\varphi$ and~$\sqrt{2}$. The $\sqrt{2}$ signal is
consistent with stochastic binary viral sharing, where the
geometric mean of engagement steps approximates~$\sqrt{2}$.

\subsection{Correct negatives}

Applied to 119,626 HYG stellar luminosities (14.5 decades),
4,307 GDP growth rates, and 100 Spotify stream counts,
\texttt{LatticeFit} returns $p > 0.15$ for all bases ---
correctly identifying these as structureless with respect
to geometric lattices.

% ── 5. Validation table ────────────────────────────────────────────
\section{Cross-domain validation}

Table~\ref{tab:validation} summarises results across 18 datasets.
Nine of eighteen yield $p < 0.05$; three yield correct negatives
at $p > 0.15$.

\begin{table}[htbp]
\centering
\small
\caption{Validation results. $n$: observations;
Dec.: $\log_{10}(x_{\max}/x_{\min})$;
$p$: log-uniform null $p$-value.
Lines above the rule: $p < 0.05$.}
\label{tab:validation}
\begin{tabular}{lrcrrrr}
\toprule
Dataset & $n$ & Dec. & Best $r$ & $d$ & RMS & $p$ \\
\midrule
Equal-tempered notes     &     13 &  0.3 & $2^{1/12}$ &  1 & $\approx 0$ & $<.001$ \\
YouTube comments (Brazil)& 15,722 &  5.2 & $\sqrt{2}$ & 12 & 0.023 & $<.001$ \\
YouTube likes (Brazil)   & 16,121 &  6.3 & $\sqrt{2}$ & 12 & 0.024 & .002 \\
Earthquake energies      &  2,054 &  3.1 & $\sqrt{2}$ &  2 & 0.011 & $<.001$ \\
Global FX rates          &  5,538 & 11.2 & $\sqrt{2}$ & 12 & 0.023 & $<.001$ \\
Intel daily volume       &  6,559 &  2.7 & $\varphi$  & 12 & 0.024 & $<.001$ \\
Intel close price        &  6,559 &  0.9 & $10$       & 12 & 0.023 & $<.001$ \\
Amazon discounted prices &  1,465 &  3.3 & $2$        & 12 & 0.022 & $<.001$ \\
Amazon actual prices     &  1,465 &  3.6 & $\varphi$  & 12 & 0.023 & $<.001$ \\
Country populations 2024 &    199 &  6.5 & $\varphi$  & 12 & 0.023 & .025 \\
Export values 2024       &    213 &  5.4 & $\varphi$  & 12 & 0.023 & .019 \\
Fungal root lengths      &     66 &  1.1 & $\varphi$  &  4 & 0.066 & .040 \\
E-commerce unit prices   &  2,000 &  2.2 & $e$        & 12 & 0.024 & .042 \\
SM fermion masses        &      9 &  5.5 & $\varphi$  &  4 & 0.069 & .074 \\
Mammal body masses       &    627 &  6.7 & $e$        &  8 & 0.035 & .083 \\
\midrule
Crude oil 1970--2026     &    675 &  2.0 & $\varphi$  & 12 & 0.024 & .221 \\
HYG stellar luminosities & 10,000 & 14.5 & $\sqrt{2}$ & 12 & 0.024 & .270 \\
GDP growth rates         &  4,307 &  4.3 & $10$       & 12 & 0.024 & .184 \\
\bottomrule
\end{tabular}
\end{table}

% ── 6. Impact ──────────────────────────────────────────────────────
\section{Impact}

\texttt{LatticeFit} provides the first domain-agnostic tool for
detecting discrete multiplicative lattice structure in empirical data.
Demonstrated applications span particle physics, geophysics, biology,
social media analytics, finance, and e-commerce. The sector-anchor
null test (patent pending, US~64/013,306) is a novel statistical
contribution with no analogue in existing software.

The publication-bundle generator directly addresses reproducibility
in lattice-structure analyses: a single command produces all
materials needed to include a rigorous lattice analysis in a
scientific paper, including a self-contained reproduction script.

% ── Declarations ───────────────────────────────────────────────────
\section*{Competing interests}
The author declares no competing interests. A US provisional patent
application (No.~64/013,306) covers the sector-anchor null test;
all other functionality is MIT licensed.

\section*{Data and code availability}
All code and example data are available at
\url{https://github.com/drmlgentry/latticefit} under the MIT License.

\section*{AI usage}
Claude (Anthropic) assisted with code development and manuscript
drafting. All scientific results were verified by the author.

% ── References ─────────────────────────────────────────────────────
\begin{thebibliography}{9}

\bibitem{GutenbergRichter1944}
B.~Gutenberg and C.~F. Richter,
Frequency of earthquakes in California,
\textit{Bull.\ Seismol.\ Soc.\ Am.}\ \textbf{34} (1944) 185.

\bibitem{Gentry2026masses}
M.~L. Gentry,
Geometric unification of flavor: masses, mixing, and CP
from the golden ratio lattice, preprint (2026).

\bibitem{scipy}
P.~Virtanen et al.,
SciPy 1.0: fundamental algorithms for scientific computing in Python,
\textit{Nature Methods}\ \textbf{17} (2020) 261.
\doi{10.1038/s41592-019-0686-2}

\bibitem{numpy}
C.~R. Harris et al.,
Array programming with NumPy,
\textit{Nature}\ \textbf{585} (2020) 357.
\doi{10.1038/s41586-020-2649-2}

\bibitem{pandas}
The Pandas Development Team,
pandas-dev/pandas, Zenodo (2024).
\doi{10.5281/zenodo.3509134}

\bibitem{matplotlib}
J.~D. Hunter,
Matplotlib: a 2D graphics environment,
\textit{Comput.\ Sci.\ Eng.}\ \textbf{9} (2007) 90.
\doi{10.1109/MCSE.2007.55}

\bibitem{lmfit}
M.~Newville et al.,
LMFIT: non-linear least-square minimization and curve-fitting for Python,
Zenodo (2014).
\doi{10.5281/zenodo.11813}

\bibitem{powerlaw}
J.~Alstott, E.~Bullmore, and D.~Plenz,
powerlaw: a Python package for analysis of heavy-tailed distributions,
\textit{PLOS ONE}\ \textbf{9} (2014) e85777.
\doi{10.1371/journal.pone.0085777}

\end{thebibliography}

\end{document}
